Profitability Prediction of Consulting Projects in the Construction Industry

### 2016-08-01 finalwriteup/draft.rmd
almost finished reading results and method

The database also excludes detailed technical information about each project. Such details are found in the preliminary drawings describing the project (square meterage, number of storeys), as well as project summary documents provided by the client to enable a quote. However, this level of information actually contradicts the purpose of the 'outside view', the intention of which is to present similar projects without being over-influenced by finer details. Given the lack of detailed information describing the physical size of projects, it was not reasonable for an algorithm to predict the numeric fee. Instead, it was more logical to predict the profitability of a project i.e. return per dollar. This measure represents the ratio of the amount invoiced to business costs, and can easily be compared between projects of different sizes.

@@

In addition to addressing cost estimation with a different kind of data set, ensemble tree methods are applied which have been minimally tested on this problem. Ensemble tree methods combine hundreds or thousands of decision trees to make predictions. A single decision tree is a non-parametric machine-learned model built by progressively determining the best binary split to partition the data according to the response variable. For further explanation of the method behind decision trees and ensemble trees, three excellent sources are @Hastie2009, @Elith2008, and @Breiman2005. To date, Linear Regression, Neural Networks, Case Based Reasoning, and Support Vector Machines (SVM's) have been used to predict cost/effort even though ensemble tree methods can perform as well as Neural Networks and generally outperform Linear Regression [@Caruana2006]. Another benefit of ensemble trees is the output types available, such as partial dependency plots and variable importance plots that provide insight into the model's calculations. This contrasts Neural Networks and SVM's, which are 'black-box' predictors. Providing insight to a user is a major benefit as past research has demonstrated issues with user trust and uptake of models, even if the models predicted well. It is anticipated that using internal data will also improve trust and create tangible meaning for the user. Blending multiple machine learning and statistical models has also not yet been applied to the cost estimation problem and may improve predictions.

@@

Chapter 4 describes the method followed throughout the research process. It includes first how the data was obtained, the lengthy cleaning process, followed by variable importance analysis, variable selection, trials of selected predictive algorithms and problem constructs. The cost estimation problem was attempted first as a regression problem predicting return per dollar, followed by a simpler binary classification problem predicting profit or loss. Once the best methods were selected, they were blended in numerous ways, using both simple averaging techniques and sophisticated machine-learning algorithms. These were compared against individual models and the best constructs were selected. Finally the impact of the algorithm on the overall profits of the case study company was analysed via decision-making scenarios. Further applications were also considered.

@@

Research in the construction industry has primarily focused on predicting the final building cost of construction. Some studies reported surveyed cost estimation practices while others tested new methods such as building information modeling (BIM) analysis, case based reasoning, and predictive algorithms. The cost estimation methods of detailed analysis (the 'inside view'), Case Based Reasoning, and predictive algorithms are reviewed in this section.

@@

Detailed analysis remains the most prevalent method for construction cost estimation despite the industry having a long history of projects running over time and budget using the same technique [@Shane2009]. Next, the method of detailed sums shall be compared to results from parametric algorithms and case based reasoning ('outside view').

@@

#### Case Based Reasoning in the Construction Industry

Case based reasoning (CBR) is a systematic method of expert judgment, where the decision maker manually compares similar projects from his or her experience [@Shepperd1996]. The idea is that in the case of a new project, similar projects are chosen, before generalisations about the data are made. This contrasts to algorithmic models that are built from the entire set of data [@Elfaki2014]. Research into the application of case based reasoning to construction cost estimation began in the 1980's [@Kim2004]. The general procedure involves:

@@

In the software industry, the main component of cost is *effort* as opposed to the cost of building materials in the construction industry. A parallel can be drawn between software development effort estimation and consultants' effort estimation in the construction industry - which was highlighted as a gap in the previous section. Plenty of research has been dedicated to the problem of effort prediction, with methods that can be categorised similarly to the construction industry: detailed analysis, algorithmic modeling, and Case Based Reasoning.

@@

#### Case Based Reasoning in the IT Industry

@@

The literature pertaining to the aim of this research project, cost prediction using internal consulting CRM data, has been extensively reviewed. Previous studies addressing the problem have generally been limited to prediction of construction costs for buildings/infrastructure and effort estimation for IT projects. The prediction models consisted of Case Based Reasoning, Linear Regression, Neural Networks and in one or two instances ensemble trees or SVM. This leaves a gap in effort estimation for consulting companies in the construction industry in combination with trialing a broader range of prediction methods.

@@

* Return per dollar as the measure of profitability

@@

Upon review of the outliers with an employee, many projects that appeared wildly profitable were actually small jobs and high profitability ratios only indicated extra profits of a few hundred dollars. It was concluded that small job outliers most likely required so little time that busy employees did not bother writing down their hours for that job. Other bigger outlier projects had unrealistically low hours logged against them which could have been the result of data entry errors or errors from when the data was transferred across databases about 4 years prior to data extraction. Based on the employee's domain knowledge and assessment of the outlier cases, reasonable cut-off values for the range of return per dollar values were revised.

@@

To create the best possible model, only important variables were selected in order to reduce the effect of noisy explanatory variables and enhance simplicity and model comprehension. Initially there were 34 explanatory variables for the single response variable, return per dollar, i.e. the chosen measure of profitability for a project. By the end of variable selection, only 11 explanatory variables remained. As stated in the previous chapter, three predictive methods were selected which output variable importance scores which differ due to their unique theoretical foundations. These were Linear Regression, Random Forests, and cForests.

@@

Linear Regression must be fed complete data sets but the case study data set only had a core set of 13 complete variables (including the response variable). If an incomplete variable was added to the core variables, the 'complete' data set would shrink to the size of the incomplete variable. If another incomplete variable was added, the complete set of data would shrink again. Therefore, to make the most of the available data, each incomplete variable was added to the core variables one at a time, where a Linear Regression model was built for each. That meant a model was made for each incomplete variable, resulting in 21 separate ANOVA models. The p-values for the F-statistic of each variable are plotted below. The core variables received p-values in all 21 models, while 'add.variable' represents p-values for the additional incomplete variable in each model.

```{r, echo=FALSE, fig.height=4.7, fig.cap = c("P-values of an ANOVA regression's F-statistics used to interpret variable importance")}

@@

ggplot(plot.pvals, aes(x = mod.vars, y = p.val)) + geom_boxplot() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
              panel.background = element_rect(fill = 'gray95')) +
        geom_hline(yintercept = 0.05, colour = 'red', linetype = 2) +
        scale_y_continuous(breaks = c(0, 0.05, .25, .5, .75, 1)) +
        labs(y = "P-value of F-statistic", x = "Independent Variables",
             title = "Summary of variable F-statistic P-values from ANOVA models")

@@

The ANOVA models indicated that the following variables have a median p-value below 0.05 and therefore significantly contribute to the rejection of the null hypothesis:

@@

The Random Forest algorithm can process data sets with missing values, however it does so by automatically imputing missing values [@randomForest]. This was not desirable in the early stage of variable selection. Instead, the completeness of the variables was assessed and it was decided that all variables with at least 2300 overlapping complete cases would be chosen. Additionally, the Random Forest algorithm cannot process categorical variables with more than 53 categories. This eliminated more variables and the resulting 'core' data set contained 15 variables and 2364 cases. Because of the Random Forest algorithm's known biases and limitations with numbers of categories, a briefer approach to variable importance analysis was taken in comparison to regression and cForests.

Each tree in a Random Forest is created with a bootstrapped training sample of the data. The random sampling means the results of each forest are slightly different. To cater for the variation in results, 10 forests were run and the mean rank of each variable was recorded. Variable importance output from a single forest is simply a ranked list of the variables where rank 1 indicates the most important variable:

@@

## Cforest

CForests were also run with the core 15 complete variables with 2364 complete cases. Incomplete variables were added to the core variables individually, each with their own separate run. Variables with an unlimited number of categories can be included in the cForest function, which is an advantage over Random Forests [@Hothorn2006; @Strobl2007; @Strobl2008]. CForests compute a variable importance ranking which is a number relating to the reduction in error the variable provides (similar method to the permutation importance in Random Forests). The variable importance rankings for the core 15 variables are as follows:

@@

ggplot(data=cfor.varimp, aes(x=var, y=imp)) +
        theme(axis.text.x=element_text(angle=45,hjust=1),
              panel.background = element_rect(fill = 'gray95')) +
        geom_bar(stat='identity') +
        labs(y = "Relative Variable Importance Value", x = "Independent Variables", title = "cForest Variable Importance Rankings of Complete Variables") +
        theme(panel.background = element_rect(fill = 'gray95'))

@@

Four incomplete variables that were added independently ranked in the top 6 most important variables in their model:

@@

The variable importance results from the three types of models: ANOVA regression, Random Forest, and cForest showed more similarities than differences. All three models ranked five variables highly:

* percent of hours by a professional
* project time span
* number of employees in project team
* total amount invoiced
* business category of client

This was not necessarily expected, particularly between ANOVA and the ensemble tree methods as ANOVA measures linear relationships in contrast to non-linear ensemble trees. Random forests favour variables that are numeric or have many categories and this held true for the analysis. The 6 highest-ranked variables were numeric with one categorical variable, Business category of client, which had 28 categories. The two lowest ranking variables had only 4 categories. It was expected that cForests would rank the categorical variables with only 4 categories more fairly. This was the case with Discipline and it received a mid-level ranking. Business category of the client and Broad business type however, were ranked similarly to Random Forests.

The qualities of the higher ranked variables were not surprising, and were generally centered around who did how much of the work internally, client characteristics, project time span, and project category. It was expected that Billing Type would play a more important role in predicting profitability as it describes the structure of the incoming fees. For this reason, and with advice from the case study business, the Billing Type was kept in the model.

For construction projects and IT software projects, the literature indicated size was the most important variable predicting effort. Size could mean size of a building or function point (level of functionality of the software) [@Shepperd1996; @Finnie1997; @Pai2013]. In this case study, profitability is the dependent variable, not total effort and the size variable plays a different role. Instead of contributing to predicting the total effort, it is describing whether the size of a project correlates with how profitable the project is. The closest estimate of size in this case study is the invoiced amount variable, which did rank as very important. The exact invoiced amount is difficult to guess before a project begins, and is part of the problem this study is addressing. Therefore, the invoiced amount was binned into categories that a manager could much more confidently choose as an estimate. The other variables indicated as important in this study  (including the number of employees on a project, client business type and percent of hours by certain types of employees) have not been strongly represented in the literature to date. This is because the variables were easily calculated from internal CRM data but would not be easily obtained externally from projects across many companies or regions.

The 11 explanatory variables from table __ were progressed to the next stage where they were used to build predictive models. The outcomes of the regression and categorical models are summarised and discussed in the following two chapters.

@@

Prediction of 'return per dollar' as a regression problem was attempted first followed by prediction of profit or loss as a binary classification problem. Linear Regression, Random Forests and Boosted Tree methods were applied to both problems with varying success.

@@

In ANOVA, categorical and continuous explanatory variables are treated differently. For categorical variables, differences in the mean values of the response variable in each group are tested. In contrast, linear relationships are examined between continuous explanatory variables and the response variable. Since linear relationships are being tested, continuous variables are assumed to be normally distributed. Therefore, where appropriate, numeric variables were normalised via logarithmic or cube root transformations. Trials of various variable interactions revealed that interactions between:

@@

were statistically significant. Two types of ANOVA models were run. The first used only a 'core' list of 6 variables that were largely complete and any relevant interaction terms.

@@

The second method was developed to overcome the problem of missing data. When all variables were combined, only 15% of the project cases had complete data and therefore the ANOVA test would eliminate 85% of the data upon execution. Imputing the missing data was not desirable at that stage due to the volume of missing categorical data and an inability to ensure the imputed data was not hampering results. The second method was a more complicated procedure that required dozens of models but allowed variables with missing data to be included. First, a project was randomly selected (a row in the data set) which would have certain variables complete. The data was then filtered for all projects that had at least the same columns complete, and an ANOVA model was built on this reduced data set. This method allowed variables with a larger proportion of missing data to be included while not attempting to analyse all variables at once.

@@

results.core = readRDS("reg_lm_core.rds")
# n.samples(data = results.core$diff)

results.A2 = readRDS("reg_lm.rds")
results.A2 <- dplyr::filter(results.A2, diff>-15)
# n.samples(data = results.A2$diff)

@@

The results of the ANOVA models were assessed using the root mean squared error (RMSE) statistic of the return per dollar predictions. These were compared against a baseline RMSE, which was the RMSE of the predicted return per dollar values against the mean return per dollar of all projects. Then, the RMSE from each model was subtracted from the RMSE using mean return per dollar. If the models were effective, the RMSE should be lower for ANOVA model predictions, so the difference would be greater than 0. Below is a histogram of this difference across 50 models run on randomly sampled 75% train and 25% test sets using core variables only (method 1 using ANOVA). Note, only `r n.samples(data = results.A2$diff)` samples (models) were required for a statistical power of 80% [@pwr].

```{r, fig.height = 4.3, fig.cap = c("Distribution of the difference in RMSE between ANOVA model predictions (built on 6 core variables) and a base line predictor that uses the mean return per dollar as its prediction. If the ANOVA model is effective, the difference should be greater than 0.")}

#plot histogram to show differences
results.core %>% ggplot(., aes(x =diff, y = ..density..))+ 
        geom_histogram() + geom_density() +
        labs(x = "Difference in RMSE (Mean RMSE - Model RMSE)", title = "Density Plot of change in RMSE for 100 ANOVA models \nCore Variables Only") +
        theme(panel.background = element_rect(fill = 'gray95'))

@@

A student t-test indicates the difference is significantly above 0 with a p-value of `r t.test(results.core$diff, alternative = "greater", mu = 0)$p.value %>% format(digits = 2)`. Therefore, the null hypothesis that the difference is less than or equal to zero can be rejected. In other words, the ANOVA model improved estimates of return per dollar over using the mean return per dollar to a statistically significant degree. However, the mean difference is only \$`r mean(results.core$diff) %>% format(digits = 1)`. Furthermore, the mean RMSE in return per dollar is \$`r mean(results.core$final.RMSE) %>% format(digits = 2)`, which is unacceptably high. Almost 50 cents is a large mean error when the value is estimating the profit from \$1 spent.

Results from 100 models using method 2, which captures incomplete variables, are shown below.

```{r, fig.height = 4.3, echo = FALSE, fig.cap = c("Distribution of the difference in RMSE between ANOVA model predictions (built on all variables) and a base line predictor that uses the mean return per dollar as its prediction. If the ANOVA model is effective, the difference should be greater than 0.")}

#plot histogram to show differences
results.A2 %>% ggplot(., aes(x =diff, y = ..density..))+ 
        geom_histogram() + geom_density() +
        labs(x = "Difference in RMSE (Mean RMSE - Model RMSE)", title = "Density Plot of change in RMSE for 50 ANOVA models \nAll Variables Sampled") +
        theme(panel.background = element_rect(fill = 'gray95'))

@@

A student t-test indicates the difference is not significantly above 0, which is clear from the plot (p-value = `r t.test(results.A2$diff, alternative = "greater", mu = 0)$p.value %>% format(digits = 3)`). This method clearly performed poorly in comparison to the core variables. A reason for this could be that the mean number of cases in the data set was `r results.core$test.nrow[1]*4` for the core variables but averaged only `r (mean(results.A2$test.nrow)*4) %>% round(0)` for the sampled set of all variables. Regression results from ANOVA models were poor - the Random Forest algorithm was trialed next.

@@

```{r load_rf_results}
results.rf.core = readRDS("reg_rf_core.rds")
# n.samples(data = results.rf.core$diff)

results.rf2 = readRDS("reg_rf.rds")
# n.samples(data = results.rf2$diff)

@@

Numeric predictions of return per dollar were attempted first with core variables that were largely complete followed by a method of sampling a mix of core and less complete variables in every run (similar to the two ANOVA methods). The Random Forest parameters were first tuned using the caret package and optimal values mtry = 5 (the number of randomly selected explanatory variables to consider at each split) and ntree = 500 (the number of decision trees to ensemble) were determined [@caret]. A density plot of 50 models of the core variables using different training/test sets is shown below. The plot was created using the same procedure used in the ANOVA models where the x-axis measures the RMSE obtained by using the Random Forest model subtracted from the RMSE using the mean of all project return per dollar values. Note, only `r n.samples(data = results.rf2$diff)` samples (models) were required for a statistical power of 80% [@pwr].

```{r, fig.height = 4.3, echo = FALSE, fig.cap = c("Distribution of the difference in RMSE between randomForest predictions (built on 6 core variables) and a base line predictor that uses the mean return per dollar as its prediction. If the ANOVA model is effective, the difference should be greater than 0.")}

#plot histogram to show differences
results.rf.core %>% ggplot(., aes(x =diff, y = ..density..))+ 
        geom_histogram() + geom_density() +
        labs(x = "Difference in RMSE (Mean RMSE - Model RMSE)", title = "Density Plot of change in RMSE for 50 randomForest models \nCore Variables Only") +
        theme(panel.background = element_rect(fill = 'gray95'))

@@

This plot shows a mean return per dollar RMSE improvement of \$`r mean(results.core$diff) %>% format(digits = 1)`, which is the same as the ANOVA model. The p-value for the null hypothesis that the difference is less than or equal to zero is `r t.test(results.rf.core$diff, alternative = "greater", mu = 0)$p.value %>% format(digits = 2)` and could be rejected. Results from 100 Random Forest models using method 2 are shown below:

```{r, fig.height = 4.3, echo = FALSE, fig.cap = c("Distribution of the difference in RMSE between randomForest predictions (built on all variables) and a base line predictor that uses the mean return per dollar as its prediction. If the ANOVA model is effective, the difference should be greater than 0.")}

#plot histogram to show differences
results.rf2 %>% ggplot(., aes(x =diff, y = ..density..))+ 
        geom_histogram() + geom_density() +
        labs(x = "Difference in RMSE (Mean RMSE - Model RMSE)", title = "Density Plot of change in RMSE for 50 randomForest models \nAll Variables Sampled") +
        theme(panel.background = element_rect(fill = 'gray95'))

>>>>

The database also excludes detailed technical information about each project. Such details are found in the preliminary drawings describing the project (square meterage, number of storeys), as well as project summary documents provided by the client to enable a quote. However, this level of information actually contradicts the purpose of the 'outside view', the intention of which is to present similar projects without being over-influenced by finer details. Given the lack of detailed information describing the physical size of projects, it was not reasonable for an algorithm to predict the numeric fee. Instead, it was more logical to predict the profitability of a project i.e. 'return per dollar'. This measure represents the ratio of the amount invoiced to business costs, and can easily be compared between projects of different sizes.

@@

In addition to addressing cost estimation with a different kind of data set, ensemble tree methods are applied which have been minimally tested on this problem. Ensemble tree methods combine hundreds or thousands of decision trees to make predictions. A single decision tree is a non-parametric machine-learned model built by progressively determining the best binary split to partition the data according to the response variable. For further explanation of the method behind decision trees and ensemble trees, three excellent sources are @Hastie2009, @Elith2008, and @Breiman2005. To date, Linear Regression, Neural Networks, Case Based Reasoning (CBR), and Support Vector Machines (SVM's) have been used to predict cost/effort even though ensemble tree methods can perform as well as Neural Networks and generally outperform Linear Regression [@Caruana2006]. Another benefit of ensemble trees is the output types available, such as partial dependency plots and variable importance plots that provide insight into the model's calculations. This contrasts Neural Networks and SVM's, which are 'black-box' predictors. Providing insight to a user is a major benefit as past research has demonstrated issues with user trust and uptake of models, even if the models predicted well. It is anticipated that using internal data will also improve trust and create tangible meaning for the user. Blending multiple machine learning and statistical models has also not yet been applied to the cost estimation problem and may improve predictions.

@@

Chapter 4 describes the method followed throughout the research process. It includes first how the data was obtained, the lengthy cleaning process, followed by variable importance analysis, variable selection, trials of selected predictive algorithms and problem constructs. The cost estimation problem was attempted first as a regression problem predicting 'return per dollar', followed by a simpler binary classification problem predicting profit or loss. Once the best methods were selected, they were blended in numerous ways, using both simple averaging techniques and sophisticated machine-learning algorithms. These were compared against individual models and the best constructs were selected. Finally the impact of the algorithm on the overall profits of the case study company was analysed via decision-making scenarios. Further applications were also considered.

@@

Research in the construction industry has primarily focused on predicting the final building cost of construction. Some studies reported surveyed cost estimation practices while others tested new methods such as building information modeling (BIM) analysis, CBR, and predictive algorithms. The cost estimation methods of detailed analysis (the 'inside view'), CBR, and predictive algorithms are reviewed in this section.

@@

Detailed analysis remains the most prevalent method for construction cost estimation despite the industry having a long history of projects running over time and budget using the same technique [@Shane2009]. Next, the method of detailed sums shall be compared to results from parametric algorithms and CBR ('outside view').

@@

#### CBR in the Construction Industry

CBR is a systematic method of expert judgment, where the decision maker manually compares similar projects from his or her experience [@Shepperd1996]. The idea is that in the case of a new project, similar projects are chosen, before generalisations about the data are made. This contrasts to algorithmic models that are built from the entire set of data [@Elfaki2014]. Research into the application of CBR to construction cost estimation began in the 1980's [@Kim2004]. The general procedure involves:

@@

In the software industry, the main component of cost is *effort* as opposed to the cost of building materials in the construction industry. A parallel can be drawn between software development effort estimation and consultants' effort estimation in the construction industry - which was highlighted as a gap in the previous section. Plenty of research has been dedicated to the problem of effort prediction, with methods that can be categorised similarly to the construction industry: detailed analysis, algorithmic modeling, and CBR.

@@

#### CBR in the IT Industry

@@

The literature pertaining to the aim of this research project, cost prediction using internal consulting CRM data, has been extensively reviewed. Previous studies addressing the problem have generally been limited to prediction of construction costs for buildings/infrastructure and effort estimation for IT projects. The prediction models consisted of CBR, Linear Regression, Neural Networks and in one or two instances ensemble trees or SVM. This leaves a gap in effort estimation for consulting companies in the construction industry in combination with trialing a broader range of prediction methods.

@@

* 'Return per dollar' as the measure of profitability

@@

Upon review of the outliers with an employee, many projects that appeared wildly profitable were actually small jobs and high profitability ratios only indicated extra profits of a few hundred dollars. It was concluded that small job outliers most likely required so little time that busy employees did not bother writing down their hours for that job. Other bigger outlier projects had unrealistically low hours logged against them which could have been the result of data entry errors or errors from when the data was transferred across databases about 4 years prior to data extraction. Based on the employee's domain knowledge and assessment of the outlier cases, reasonable cut-off values for the range of 'return per dollar' values were revised.

@@

To create a high-performing model, only important variables were selected in order to reduce the effect of noisy explanatory variables and enhance simplicity and model comprehension. Initially there were 34 explanatory variables for the single response variable, 'return per dollar'. By the end of variable selection, only 11 remained. As stated in the previous chapter, three predictive methods were selected that output variable importance scores. These were Linear Regression, Random Forests, and cForests, which have unique theoretical foundations.

@@

Linear Regression must be fed complete data sets but the case study data had a core set of only 13 complete variables (including the response variable). If an incomplete variable was added to the core variables, the 'complete' data set would shrink to the complete rows of the incomplete variable. If another incomplete variable was added, the complete set of data would shrink again. Therefore, to make the most of the available data, each incomplete variable was added to the core variables one at a time, where a Linear Regression model was built for each. That meant a model was made for each incomplete variable, resulting in 21 separate ANOVA models. The p-values for the F-statistic of each variable are plotted below. The core variables received p-values in all 21 models, while 'add.variable' represents the p-values for the unique additional incomplete variable in each model.

```{r, echo=FALSE, fig.height=4.7, fig.cap = c("P-values of ANOVA regressions' F-statistics which were used to interpret variable importance")}

@@

ggplot(plot.pvals, aes(x = mod.vars, y = p.val)) + geom_boxplot() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
              panel.background = element_rect(fill = 'gray95')) +
        geom_hline(yintercept = 0.05, colour = 'red', linetype = 2) +
        scale_y_continuous(breaks = c(0, 0.05, .25, .5, .75, 1)) +
        labs(y = "P-Value of F-Statistic", x = "Independent Variables",
             title = "Summary of Variable F-Statistic P-Values from ANOVA Models")

@@

The ANOVA models indicated that the following variables have a median p-value below 0.05 and therefore significantly contribute to the rejection of the regression null hypothesis:

@@

The Random Forest algorithm can process data sets with missing values, however it does so by automatically imputing those missing values [@randomForest]. This was not desirable in the early stage of variable selection. Instead, the completeness of the variables was assessed and it was decided that all variables with at least 2300 overlapping complete cases would be chosen. Additionally, the Random Forest algorithm cannot process categorical variables with more than 53 categories. This eliminated more variables and the resulting 'core' data set contained 15 variables and 2364 complete cases. Because of the Random Forest algorithm's known biases and limitations with numbers of categories, a briefer approach to variable importance analysis was taken in comparison to Regression and cForests.

As explained in the Literature Review, each tree in a Random Forest is created with a bootstrapped training sample of the data. The random sampling means the results of each forest are slightly different. To cater for the variation in results, 10 forests were run and the mean rank of each variable was recorded. Variable importance output from a single forest is simply a ranked list of the variables where rank 1 indicates the most important variable:

@@

## CForest

CForests were also run using the core 15 complete variables with 2364 complete cases. Incomplete variables were added to the core variables individually, each with their own separate run. Variables with an unlimited number of categories can be included in the cForest function, which is an advantage over Random Forests [@Hothorn2006; @Strobl2007; @Strobl2008]. CForests compute a variable importance ranking which is a score relating to the reduction in error that the variable provides (a similar method to the permutation importance in Random Forests). The variable importance rankings for the core 15 variables were as follows:

@@

ggplot(data=cfor.varimp, aes(x=var, y=imp)) +
        theme(axis.text.x=element_text(angle=45,hjust=1),
              panel.background = element_rect(fill = 'gray95')) +
        geom_bar(stat='identity') +
        labs(y = "Relative Variable Importance Value", x = "Independent Variables", title = "CForest Variable Importance Rankings of Complete Variables") +
        theme(panel.background = element_rect(fill = 'gray95'))

@@

Four incomplete variables that were added individually ranked in the top 6 most important variables in their model:

@@

The variable importance results from the three types of models: ANOVA Regression, Random Forest, and cForest showed more similarities than differences. All three models ranked five variables highly:

* Percent of hours by a professional
* Project time span
* Number of employees in project team
* Total amount invoiced
* Business category of client

This was not necessarily expected, particularly between ANOVA and the ensemble tree methods, because ANOVA measures linear relationships in contrast to non-linear ensemble trees.

Random Forests favour variables that are numeric or have many categories and this held true for the analysis. Its 6 highest-ranked variables were numeric with one categorical variable, Business category of the client, which had 28 categories. The two lowest ranking variables had only 4 categories. It was expected that cForests would rank the categorical variables with only 4 categories more fairly. This was the case with Discipline and it received a mid-level ranking. Business category of the client and Broad business type however, were ranked similarly to Random Forests.

The overall qualities of the higher ranked variables were not surprising, and were generally centered around who did how much of the work internally, client characteristics, project time span, and project category. It was expected that Billing Type would play a more important role in predicting profitability as it describes the structure of the incoming fees. For this reason, and with advice from the case study business, the Billing Type was kept in the model.

For IT software projects, the literature indicated size was the most important variable predicting effort, and in Construction projects size was also often an explanatory variable. Size could mean size of a building or function point (level of functionality of the software) [@Shepperd1996; @Finnie1997; @Pai2013]. In this case study, profitability is the dependent variable, not total effort and the size variable plays a different role. Instead of predicting total effort, it is describing whether the size of a project correlates with how profitable the project is. The closest estimate of size in this case study is the invoiced amount variable, which did rank as very important. The exact invoiced amount is difficult to guess before a project begins, and is part of the problem this study is addressing. Therefore, the invoiced amount was binned into categories that a manager could much more confidently choose as an estimate. The other variables indicated as important in this study (including the number of employees on a project, client business type and percent of hours by certain types of employees) have not been strongly represented in the literature to date. This is because the variables were easily calculated from internal CRM data but would not be easily obtained from an external standpoint for projects across many companies or regions.

The 11 explanatory variables from table __ were progressed to the next stage where they were used to build the first set of predictive models. The outcomes of the regression and categorical models are summarised and discussed in the next two chapters.

@@

Prediction of 'return per dollar' as a regression problem was attempted first followed by prediction of profit or loss as a binary classification problem. Linear Regression, Random Forests and Boosted Tree methods were applied to both problems.

@@

In ANOVA, categorical and continuous explanatory variables are treated differently. For categorical variables, differences in the mean values and variances of the response variable in each group are tested. In contrast, linear relationships are examined between continuous explanatory variables and the response variable. The continuous variables were normalised to kerb the effects of possible high leverage points, and logarithmic or cube root transformations were used for this. Trials of various variable interactions revealed that interactions between:

@@

were statistically significant. Two types of ANOVA models were built. The first used only a 'core' list of 6 variables that were largely complete along with relevant interaction terms:

@@

The second method was developed to overcome the problem of missing data. When all variables were combined, only 15% of the project cases had complete data and therefore the ANOVA test would eliminate 85% of the data upon execution. Imputing the missing data was not desirable at that stage due to the volume of missing categorical data and an inability to ensure the imputed data did not adversely influence results. The second method was a complicated procedure that required dozens of models but allowed variables with missing data to be included. First, a project was randomly selected (a row in the data set) which would have certain variables complete. The data was then filtered for all projects that had at least the same columns complete, and an ANOVA model was built on this reduced data set. This method allowed variables with a larger proportion of missing data to be included while not attempting to analyse all variables at once.

@@

results.core = readRDS("report_data/reg_lm_core.rds")
# n.samples(data = results.core$diff)

results.A2 = readRDS("report_data/reg_lm.rds")
results.A2 <- dplyr::filter(results.A2, diff>-15)
# n.samples(data = results.A2$diff)

@@

The results of the ANOVA models were assessed using the RMSE statistic of the 'return per dollar' predictions. These were compared against a baseline RMSE, which was the RMSE of the predicted 'return per dollar' values less the mean 'return per dollar' of all projects.

$RMSE_b = \sqrt{\frac{\sum_{t = 1}^{n} (\widehat{y}_t - \bar y)^2}{n}}$

RMSE_b = baseline root mean squared error
y = predicted 'return per dollar'
y = mean 'return per dollar' for al projects

$$
\begin{aligned}
RMSE_b = \textrm{baseline root mean squared error} \\
n &= \textrm{number of data points} \\
\widehat{y} &= \textrm{predicted values of the response variable} \\
\bar y &= \textrm{mean 'return per dollar' for all projects} \\
\end{aligned}
$$

Then, the RMSE from each model was subtracted from the baseline RMSE:

$\Delta\ RMSE = RMSE_b - RMSE$

If the models were effective, the RMSE should be lower for ANOVA model predictions, so the difference would be greater than 0. Below is a histogram of this difference across 50 models run on randomly sampled 75% train and 25% test sets using core variables only (method 1 ANOVA). Note, only `r n.samples(data = results.A2$diff)` samples (models) were required for a statistical power of 80% [@pwr].

```{r, fig.height = 4.3, fig.cap = c("Distribution of the difference in RMSE between ANOVA model predictions (built on 6 core variables) and a baseline predictor that uses the mean 'return per dollar' as its prediction. If the ANOVA model was effective, the difference should be greater than 0.")}

#plot histogram to show differences

results.core %>% ggplot(., aes(x =diff, y = ..density..))+ 
        geom_histogram() + geom_density() +
        labs(x = expression(Delta~RMSE), 
             y = "Density", title = expression(atop(paste("Density Plot of")~Delta~RMSE~paste("for 50 ANOVA Models"),Core~Variables~Only))) +
        theme(panel.background = element_rect(fill = 'gray95'))

@@

A student t-test indicated $\Delta RMSE$ was significantly above 0 with a p-value of `r t.test(results.core$diff, alternative = "greater", mu = 0)$p.value %>% format(digits = 2)`. Therefore, the null hypothesis that the difference is less than or equal to zero can be rejected. In other words, the ANOVA model improved estimates of 'return per dollar' over using the mean 'return per dollar' to a statistically significant degree. However, the mean difference was only \$`r mean(results.core$diff) %>% format(digits = 1)`. Furthermore, the mean RMSE in 'return per dollar' was \$`r mean(results.core$final.RMSE) %>% format(digits = 2)`, which is unacceptably high. Almost 50 cents is a large error when the value is estimating the profit after spending \$1.

Results from 100 models using method 2, which captures incomplete variables, are shown below:

```{r, fig.height = 4.3, echo = FALSE, fig.cap = c("Distribution of the difference in RMSE between ANOVA model predictions (built on all variables) and a baseline predictor that uses the mean 'return per dollar' as its prediction. If the ANOVA model is effective, the difference should be greater than 0.")}

#plot histogram to show differences
results.A2 %>% ggplot(., aes(x =diff, y = ..density..))+ 
        geom_histogram() + geom_density() +
        labs(x = expression(Delta~RMSE), 
             y = "Density", title = expression(atop(paste("Density Plot of")~Delta~RMSE~paste("for 100 ANOVA Models"),All~Variables~Sampled))) +
        theme(panel.background = element_rect(fill = 'gray95'))

@@

A student t-test indicated the difference was not significantly above 0, which is clear from the plot (p-value = `r t.test(results.A2$diff, alternative = "greater", mu = 0)$p.value %>% format(digits = 3)`). This method clearly performed poorly in comparison to the core variables. A reason for this could be that the mean number of cases in the data set was `r results.core$test.nrow[1]*4` for the core variables but averaged only `r (mean(results.A2$test.nrow)*4) %>% round(0)` for the sampled set of all variables. Overall, the Regression results from Linear ANOVA models were poor and the Random Forest algorithm was trialed next.

@@

```{r load_rf_results}
results.rf.core = readRDS("report_data/reg_rf_core.rds")
# n.samples(data = results.rf.core$diff)

results.rf2 = readRDS("report_data/reg_rf.rds")
# n.samples(data = results.rf2$diff)

@@

Numeric predictions of 'return per dollar' were attempted first with core variables that were largely complete followed by a method of sampling a mix of core and less complete variables in every run (similar to the two ANOVA methods). The Random Forest parameters were first tuned using the caret package and the optimal values mtry = 5 (the number of randomly selected explanatory variables to consider at each split) and ntree = 500 (the number of decision trees to ensemble) were determined [@caret]. A density plot of 50 models of the core variables using different training/test sets is shown below. The plot was created using the same procedure used in the ANOVA models where the x-axis measures $\Delta RMSE$. Note, only `r n.samples(data = results.rf2$diff)` samples (models) were required for a statistical power of 80% [@pwr].

```{r, fig.height = 4.3, echo = FALSE, fig.cap = c("Distribution of the difference in RMSE between randomForest predictions (built on 6 core variables) and a baseline predictor that uses the mean 'return per dollar' as its prediction. If the Random Forest model was effective, the difference would be greater than 0.")}

#plot histogram to show differences
results.rf.core %>% ggplot(., aes(x =diff, y = ..density..))+ 
        geom_histogram() + geom_density() +
        labs(x = expression(Delta~RMSE), 
             y = "Density", title = expression(atop(paste("Density Plot of")~Delta~RMSE~paste("for 50 Random Forest Models"),Core~Variables~Only))) +
        theme(panel.background = element_rect(fill = 'gray95'))

@@

This plot shows a mean 'return per dollar' RMSE improvement of \$`r mean(results.core$diff) %>% format(digits = 1)`, which is the same as the ANOVA model. The p-value for the null hypothesis that the difference is less than or equal to zero is `r t.test(results.rf.core$diff, alternative = "greater", mu = 0)$p.value %>% format(digits = 2)` and could be rejected. Results from 100 Random Forest models using method 2 are shown below:

```{r, fig.height = 4.3, echo = FALSE, fig.cap = c("Distribution of the difference in RMSE between randomForest predictions (built on all variables) and a baseline predictor that uses the mean 'return per dollar' as its prediction. If the Random Forest model is effective, the difference should be greater than 0.")}
